---
title: "RNA-seq analysis of gliosis in rd1 and rd10 models"
author: "Alejandro Gombau García"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    toc: true
header-includes:
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

```{r message=FALSE, warning=FALSE, include=FALSE}

library(GEOquery)
library(Matrix)
library(Seurat)
library(DESeq2)
library(clusterProfiler)
library(org.Mm.eg.db)
library(EnhancedVolcano)
library(patchwork)

```

RNA-Seq permits the precise and large-scale determination of RNA sequences present within a sample, facilitating the quantification of transcript abundance and the discovery of novel variants or isoforms. It is primarily employed to examine gene expression patterns across different conditions (e.g., studying retinopathies in disease models versus wild-type strains), identify differentially expressed genes, characterize tissue transcriptomes, and detect non-coding RNAs or gene fusions.

Retinal degenerative processes often involve a sequence where chronic loss of ganglion cells and axons induces severe gliosis. This glial environment exacerbates axonal atrophy and neuronal degeneration.

Utilizing RNA-Seq, this analysis investigates these processes using datasets retrieved from GEO (Gene Expression Omnibus). The goal is to elucidate gene expression changes associated with glial activation, inflammation, and the downregulation of neuroprotective molecules, thereby identifying the associated molecular signatures.

The central question addressed through this pipeline is: **Is it feasible to search for IHC markers on a standard laptop using public RNA-seq data?** By conducting this computational evaluation, the analytical pipeline is calibrated for future investigations involving other retinopathy models that may share convergent phenotypes.

# Data Preparation

The dataset originates from the Gene Expression Omnibus (GEO), a public repository for high-throughput gene expression data maintained by NCBI. Single-cell RNA-seq datasets from two recent studies are utilized:

-   Simon CJ, Khabou H, Chaffiol A, Rucli M et al. [Reactivating the phototransduction cascade with a mutation agnostic gene therapy preserves vision in rod-cone dystrophies.](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE286529) iScience 2025. Data from rd10 and rd1 models are obtained from this study.
-   Heng JS, Hackett SF, Stein-O'Brien GL, Winer BL et al. [Comprehensive analysis of a mouse model of spontaneous uveoretinitis using single-cell RNA sequencing.](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE132229) Proc Natl Acad Sci U S A 2019. Data from the wild-type (C57) model are obtained from this study.

Datasets for *rd16*, *rd12*, and *rd8* mouse strains were unavailable, presenting a potential avenue for future in-house studies.

Initially, data from various retinal samples are downloaded from GEO. Following quality control to remove low-quality instances, cells from each sample are aggregated into a pseudo-bulk representation. This approach simplifies the single-cell RNA-seq data into a format akin to bulk RNA-seq, representing each sample as a single list of genes and their associated read counts, thereby standardizing the input for differential expression analysis.

Extensive code sections dedicated to data downloading, formatting, quality control, and pseudo-bulk construction execute in the background.

```{r include=FALSE}

# Define a vector with sample IDs
gsm_ids <- c(
  "GSM8728964", # Retina rd1 mouse sample 1
  "GSM8728965", # Retina rd1 mouse sample 2
  "GSM8728966", # Retina rd1 mouse sample 3
  "GSM8728967", # Retina rd10 mouse sample 1
  "GSM8728968", # Retina rd10 mouse sample 2
  "GSM3854512"  # JH21_WT_retina_scRNAseq_R1
)

# Download metadata for each GSM without quantitative expression series
gsm_list <- lapply(gsm_ids, function(id) {
    getGEO(id, GSEMatrix = FALSE)
})

```

```{r warning=FALSE, include=FALSE}
# Create a folder for each GSM and download supplementary files
for (gsm in gsm_ids) {
  dir.create(paste0(gsm, "_data"), showWarnings = FALSE)
  getGEOSuppFiles(gsm, makeDirectory = FALSE, baseDir = paste0(gsm, "_data"))
}
```

```{r include=FALSE}
# Verify contents of all folders
for (gsm in gsm_ids) {
  cat("Files in", gsm, "_data:\n")
  print(list.files(paste0(gsm, "_data")))
  cat("\n")
}
```

```{r include=FALSE}
# Define IDs and folders containing files for each GSM
gsm_ids <- c(
  "GSM8728964", 
  "GSM8728965", 
  "GSM8728966", 
  "GSM8728967", 
  "GSM8728968", 
  "GSM3854512"
)
# Construct folder names for each GSM
carpetas <- paste0(gsm_ids, "_data")

# Prepare an empty list to store matrices
lista_expr <- vector("list", length(gsm_ids))
names(lista_expr) <- gsm_ids

# Loop to read each sample
for (i in seq_along(gsm_ids)) {
  gsm   <- gsm_ids[i]
  dir   <- carpetas[i]
  
  # Identify filenames within the folder
  ficheros <- list.files(dir, full.names = TRUE)
  
  # Read barcodes 
  fn_barcodes <- ficheros[grep("_barcodes.tsv.gz$", ficheros)]
  barcodes    <- read.delim(fn_barcodes, header = FALSE, stringsAsFactors = FALSE)
  
  # Read features or genes 
  fn_features <- ficheros[grep("_features.tsv.gz$|_genes.tsv.gz$", ficheros)]
  features    <- read.delim(fn_features, header = FALSE, stringsAsFactors = FALSE)
  
  # Read the sparse matrix
  fn_matrix <- ficheros[grep("_matrix.mtx.gz$", ficheros)]
  expr_mat  <- readMM(fn_matrix)
  
  # Assign row and column names
  rownames(expr_mat) <- features$V2
  colnames(expr_mat) <- barcodes$V1
  
  # Save the matrix in the list
  lista_expr[[gsm]] <- expr_mat
  
  # Show a brief summary of dimensions
  cat("->", gsm, ": number of genes =", nrow(expr_mat),
      ", number of cells =", ncol(expr_mat), "\n")
  
  # Free temporary variables
  rm(expr_mat, barcodes, features)
  gc()
}

```

```{r include=FALSE}
# Make gene names unique for each sample
for (gsm in names(lista_expr)) {
  genes_orig      <- rownames(lista_expr[[gsm]])
  genes_unicos    <- make.unique(genes_orig)
  rownames(lista_expr[[gsm]]) <- genes_unicos
  
  rm(genes_orig, genes_unicos)
}
gc()

# Verify duplicated genes in each sample
for (gsm in names(lista_expr)) {
  n_dup <- sum(duplicated(rownames(lista_expr[[gsm]])))
  cat("->", gsm, ": duplicated genes =", n_dup, "\n")
  
  rm(n_dup)
}
gc()

```

```{r include=FALSE}
# Initialize an empty list to store Seurat objects
lista_seurat <- vector("list", length(lista_expr))
names(lista_seurat) <- names(lista_expr)

# Loop to create each Seurat object
for (gsm in names(lista_expr)) {
  lista_seurat[[gsm]] <- CreateSeuratObject(
    counts  = lista_expr[[gsm]],
    project = gsm
  )
  cat("-> Created Seurat for", gsm, ":",
      nrow(lista_seurat[[gsm]]), "genes x",
      ncol(lista_seurat[[gsm]]), "cells\n")
}

# Remove the original list of matrices and force garbage collection to manage RAM limit
rm(lista_expr)
gc()

```

```{r include=FALSE}
for (gsm in names(lista_seurat)) {
  seu <- lista_seurat[[gsm]]
  
  # Calculate percent.mt if it does not exist
  if (!"percent.mt" %in% colnames(seu@meta.data)) {
    seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^mt-")
    lista_seurat[[gsm]] <- seu  # save the updated object
  }
  
  # Print the summary for that sample
  cat(">>>", gsm, "<<<\n")
  cat("Summary of nFeature_RNA:\n")
  print(summary(seu@meta.data$nFeature_RNA))
  cat("Summary of percent.mt:\n")
  print(summary(seu@meta.data$percent.mt))
  cat("\n-------------------------\n\n")
  
  rm(seu)
}
gc()

```

```{r include=FALSE}
# Apply filtering to each Seurat object and display final cell count
for (gsm in names(lista_seurat)) {
  seu <- lista_seurat[[gsm]]
  
  # Filter objects based on transcripts and mitochondrial RNA fraction
  seu_filtrado <- subset(
    seu,
    subset = nFeature_RNA > 200 & percent.mt < 10
  )
  
  # Save the filtered Seurat object in the list
  lista_seurat[[gsm]] <- seu_filtrado
  
  # Print the number of remaining cells
  cat("->", gsm, ": cells after QC =", ncol(seu_filtrado), "\n")
  
  rm(seu, seu_filtrado)
}
gc()

```

```{r include=FALSE}
# Extract the vector of present genes in each filtered sample using GetAssayData
lista_genes_por_muestra <- lapply(lista_seurat, function(s) {
  counts_mat <- GetAssayData(s, layer = "counts")
  rownames(counts_mat)
})

# Calculate the intersection to extract common genes
genes_comunes_QC <- Reduce(intersect, lista_genes_por_muestra)

# Print the number of common genes
length(genes_comunes_QC)

rm(lista_genes_por_muestra)
gc()

```

```{r include=FALSE}
# Create empty matrix: rows = common genes, columns = each sample
pseudo_bulk_mat_QC <- matrix(
  0,
  nrow = length(genes_comunes_QC),
  ncol = length(lista_seurat),
  dimnames = list(genes_comunes_QC, names(lista_seurat))
)

# Traverse each sample and aggregate gene transcript counts
for (gsm in names(lista_seurat)) {
  # Extract the counts matrix from the Seurat object
  counts_qc <- GetAssayData(lista_seurat[[gsm]], layer = "counts")
  
  # Select only the common genes
  submat    <- counts_qc[genes_comunes_QC, , drop = FALSE]
  
  # Compute row sums and register in the corresponding pseudo-bulk column
  pseudo_bulk_mat_QC[, gsm] <- Matrix::rowSums(submat)
  
  rm(counts_qc, submat)
}
gc()

# Convert the structural matrix format to data.frame
pseudo_bulk_df_QC <- as.data.frame(pseudo_bulk_mat_QC)

# Print dimensions and dataframe sample
cat("Dimensions of pseudo-bulk after QC:", dim(pseudo_bulk_df_QC), "\n\n")
head(pseudo_bulk_df_QC)

# Discard large redundant objects
rm(pseudo_bulk_mat_QC, lista_seurat)
gc()

```

```{r include=FALSE}
# Install and load DESeq2 if necessary
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("DESeq2")

# Create a data.frame with the structural sample information
colData <- data.frame(
  sample     = colnames(pseudo_bulk_df_QC),
  condition  = c(
    "rd1", "rd1", "rd1",   # GSM8728964, 8965, 8966 correspond to rd1
    "rd10","rd10",         # GSM8728967, 8968 correspond to rd10
    "wt"                  # GSM3854512 corresponds to C57 WT
  ),
  row.names = colnames(pseudo_bulk_df_QC),
  stringsAsFactors = FALSE
)
colData

```

```{r include=FALSE}
# Construct the colData data.frame properly formatted
colData <- data.frame(
  condition = c(
    "rd1", "rd1", "rd1",   
    "rd10","rd10",         
    "wt"                   
  ),
  row.names = colnames(pseudo_bulk_df_QC),
  stringsAsFactors = FALSE
)

# Verify dimensional parameters and content
dim(colData)
colData

```

# Differential Expression

```{r include=FALSE}
# Create the DESeqDataSet architecture utilizing the count matrix and colData
dds <- DESeqDataSetFromMatrix(
  countData = pseudo_bulk_df_QC,
  colData   = colData,
  design    = ~ condition
)

# Render class overview
dds

# Execute the core differential expression analysis routine
dds <- DESeq(dds)
```

## Contrasts Table

The summarized count matrix generated longitudinally is leveraged alongside `DESeq2` to execute differential comparisons. For individual genes, `DESeq2` derives a log2 fold change (logFC)—delineating proportional regulation ratios—and an adjusted p-value (`padj`) defining structural statistical significance. Only variables meeting standard coverage limits (`baseMean` \>= 6) and satisfying padj \< 0.1 criteria are assigned as structurally differentially expressed.

```{r echo=FALSE}
# Extract differential outputs across contrast architectures
res_rd1_vs_wt   <- results(dds, contrast = c("condition", "rd1",  "wt"))
res_rd10_vs_wt  <- results(dds, contrast = c("condition", "rd10", "wt"))
res_rd1_vs_rd10 <- results(dds, contrast = c("condition", "rd1",  "rd10"))

# Auxiliary function calculating Total, Up, Down, Base ratios 
resumen_contraste_pct <- function(res_obj) {
  # Coerce to structural data.frame
  df <- as.data.frame(res_obj)
  # Filter underlying genes possessing mean counts >= 6
  df_filt <- df[!is.na(df$padj) & df$baseMean >= 6, ]
  
  total   <- nrow(df_filt)
  up      <- sum(df_filt$padj < 0.1 & df_filt$log2FoldChange >  0, na.rm = TRUE)
  down    <- sum(df_filt$padj < 0.1 & df_filt$log2FoldChange <  0, na.rm = TRUE)
  pct_up   <- up    / total * 100
  pct_down <- down  / total * 100
  
  return(c(
    Total   = total,
    Up      = up,
    Down    = down,
    PctUp   = round(pct_up, 2),
    PctDown = round(pct_down, 2)
  ))
}

# Consolidate output ratio framework
tabla_resumen_pct <- rbind(
  `rd1_vs_wt`   = resumen_contraste_pct(res_rd1_vs_wt),
  `rd10_vs_wt`  = resumen_contraste_pct(res_rd10_vs_wt),
  `rd1_vs_rd10` = resumen_contraste_pct(res_rd1_vs_rd10)
)

# Output summary layout
tabla_resumen_pct

```

Results delineate that in the rd1 vs WT contrast, out of 17,466 analyzed genes, 2,716 (approx. 15.6%) present upregulated profiles in rd1, while 2,862 (approx. 16.4%) manifest downregulation. In the rd10 vs WT contrast, of 16,592 genes, 2,657 (approx. 16.0%) are upregulated in rd10 and 2,770 (approx. 16.7%) are downregulated.

Evaluating the comparative rd1 vs rd10 schema, out of 17,490 total targets, only 292 (1.7%) display significant elevation in rd1 and 407 (approx. 2.3%) present elevation in rd10. Both experimental retinopathic vectors profoundly deviate from the initial WT baseline, retaining remarkable structural alignment among themselves.

## Shrinkage Parameterization

The shrinkage method in `DESeq2` adjusts log2 fold change estimates to provide more stable results, particularly for genes with low read counts. Because low-count genes often exhibit high variability, shrinkage reduces false positives by pulling highly uncertain fold changes closer to zero.

We extracted the expression values for established markers of glial reactivity, including astrocytes, Müller glia, and microglia:

"Gfap", "Vim", "S100b", "Aldh1l1", "Rlbp1", "Glul", "Slc1a3", "C1qa", "C1qb", "Trem2", "Aif1"

The results detail the normalized mean read count (`baseMean`), log2 fold change (`log2FoldChange`), standard error (`lfcSE`), p-value, and adjusted p-value (`padj`) for the rd1 vs WT and rd10 vs WT comparisons.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Process adaptive shrinkage for rd1 vs wt 
resLFC_rd1_vs_wt <- lfcShrink(
  dds,
  contrast = c("condition", "rd1", "wt"),
  type     = "ashr"
)

# Process adaptive shrinkage for rd10 vs wt
resLFC_rd10_vs_wt <- lfcShrink(
  dds,
  contrast = c("condition", "rd10", "wt"),
  type     = "ashr"
)

# Process adaptive shrinkage for rd1 vs rd10 
resLFC_rd1_vs_rd10 <- lfcShrink(
  dds,
  contrast = c("condition", "rd1", "rd10"),
  type     = "ashr"
)

# Declare comprehensive gliosis markers
genes_gliosis <- c(
  "Gfap", "Vim", "S100b", "Aldh1l1",
  "Rlbp1", "Glul", "Slc1a3",
  "C1qa", "C1qb", "Trem2", "Aif1"
)

# Coerce derived structural data elements
tablaLFC_rd1  <- as.data.frame(resLFC_rd1_vs_wt)
tablaLFC_rd10 <- as.data.frame(resLFC_rd10_vs_wt)

# Filter markers parallel defined profiles
tabla_gliosis_rd1  <- tablaLFC_rd1[rownames(tablaLFC_rd1) %in% genes_gliosis, ]
tabla_gliosis_rd10 <- tablaLFC_rd10[rownames(tablaLFC_rd10) %in% genes_gliosis, ]

# Output resultant data structures
head(tabla_gliosis_rd1)
head(tabla_gliosis_rd10)


```

In the rd1 vs WT comparison, nearly all glial markers are significantly upregulated. This points to clear astrogliosis, Müller cell reactivity (Glul, Rlbp1, Slc1a3), and an increase in microglial markers (C1qa, C1qb, Aif1). The rd10 vs WT contrast shows a very similar pattern, although the magnitude of these changes is slightly lower. Overall, these results confirm that gliosis-associated genes are elevated in both models, supporting the role of glial reactivity in retinal degeneration.

# Functional Enrichment Analysis

Functional enrichment analysis identifies biological processes that are overrepresented within a list of differentially expressed genes. Rather than analyzing genes individually, this approach groups them to reveal broader functional trends. Comparing the upregulated genes against the Gene Ontology (Biological Process) database highlights a clear association with gliogenesis and cytoskeletal reorganization.

```{r echo=FALSE}
# Coerce filtered responses
df_rd1_vs_wt  <- as.data.frame(resLFC_rd1_vs_wt)
df_rd10_vs_wt <- as.data.frame(resLFC_rd10_vs_wt)

# Filter parameters enforcing padj < 0.1 criteria coupled with proportional baseline increases
genes_up_rd1  <- rownames(df_rd1_vs_wt)[df_rd1_vs_wt$padj < 0.1 & df_rd1_vs_wt$log2FoldChange >  0]
genes_up_rd10 <- rownames(df_rd10_vs_wt)[df_rd10_vs_wt$padj < 0.1 & df_rd10_vs_wt$log2FoldChange >  0]

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

entrez_rd1 <- bitr(
  genes_up_rd1,
  fromType   = "SYMBOL",
  toType     = "ENTREZID",
  OrgDb      = org.Mm.eg.db
)$ENTREZID

entrez_rd10 <- bitr(
  genes_up_rd10,
  fromType   = "SYMBOL",
  toType     = "ENTREZID",
  OrgDb      = org.Mm.eg.db
)$ENTREZID
```

```{r echo=FALSE}
# Perform GO Biological Process statistical derivation representing rd1
ego_rd1 <- enrichGO(
  gene          = entrez_rd1,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE   # Transpose internal ENTREZID configurations to standard notation
)

# Perform GO derivation representing rd10
ego_rd10 <- enrichGO(
  gene          = entrez_rd10,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# Subset high priority integrated routes
top10_rd1  <- as.data.frame(ego_rd1)[1:10,  ]
top10_rd10 <- as.data.frame(ego_rd10)[1:10, ]

# Output derivations
cat("=== Top 10 GO BP in rd1 vs WT ===\n")
print(top10_rd1[, c("ID", "Description", "GeneRatio", "p.adjust")])
cat("\n=== Top 10 GO BP in rd10 vs WT ===\n")
print(top10_rd10[, c("ID", "Description", "GeneRatio", "p.adjust")])

```

In both models, functional enrichment analysis highlights gliogenesis as a highly significant pathway. Additionally, processes related to the regulation of neurogenesis, cytoskeleton organization, and cell adhesion show significant divergence from healthy controls. These pathway shifts suggest widespread structural reorganization and strong glial reactivity in the degenerating retina.

# Volcano Plot

Volcano plots visually map transcriptomic differences between experimental groups. The horizontal axis represents the magnitude of gene expression changes, while the vertical axis indicates the statistical significance of those alterations.

```{r echo=FALSE}
# Map explicit representations rd1 vs WT
df_volc_rd1 <- as.data.frame(resLFC_rd1_vs_wt)
df_volc_rd1$gene <- rownames(df_volc_rd1)

# Map explicit representations rd10 vs WT
df_volc_rd10 <- as.data.frame(resLFC_rd10_vs_wt)
df_volc_rd10$gene <- rownames(df_volc_rd10)

```

```{r warning=FALSE, include=FALSE}
p_rd1 <- EnhancedVolcano(
  df_volc_rd1,
  lab            = df_volc_rd1$gene,
  x              = "log2FoldChange",
  y              = "padj",
  xlab           = bquote(~Log[2]~ "(Fold Change)"),
  ylab           = bquote(~-Log[10]~ "(adj P-value)"),
  pCutoff        = 0.05,
  FCcutoff       = 1,
  pointSize      = 2.0,
  labSize        = 3.0,
  title          = "rd1 vs WT",
  subtitle       = "padj < 0.05, |log2FC| > 1",
  caption        = "Upregulated targets represented in red",
  drawConnectors = TRUE,
  max.overlaps   = 30
)

p_rd10 <- EnhancedVolcano(
  df_volc_rd10,
  lab            = df_volc_rd10$gene,
  x              = "log2FoldChange",
  y              = "padj",
  xlab           = bquote(~Log[2]~ "(Fold Change)"),
  ylab           = bquote(~-Log[10]~ "(adj P-value)"),
  pCutoff        = 0.05,
  FCcutoff       = 1,
  pointSize      = 2.0,
  labSize        = 3.0,
  title          = "rd10 vs WT",
  subtitle       = "padj < 0.05, |log2FC| > 1",
  caption        = "Upregulated targets represented in red",
  drawConnectors = TRUE,
  max.overlaps   = 30
)

p_rd1 + p_rd10 + plot_layout(ncol = 2, guides = "collect") & theme(legend.position = "bottom")
```

![Volcano plot of differentially expressed genes in rd1 and rd10 retinopathy models compared to wild-type (WT) controls. The x-axis represents the log2 fold change of gene expression, and the y-axis indicates the statistical significance (-log10 of the adjusted p-value). Each point represents a single gene. The vertical dashed lines indicate the fold change threshold (\|log2FC\| \> 1), and the horizontal dashed line marks the significance threshold (padj \< 0.05). Genes highlighted in red represent upregulated targets meeting both criteria. Analysis performed using DESeq2 with shrinkage adjustment.](000014.png)

The generated plot can serve as a preliminary map to help index potential shifts between the wild-type baseline and the degenerative models. Variables positioned in the upper right quadrant might indicate components with higher relative expression. The presence of established structural indicators in these regions, if observed, could lend support to the model's validity. This approach may help prioritize potential biological targets, offering a possible avenue to identify specific regulatory species for further validation using standard IHC techniques.

# Conclusions

This pipeline serves as a proof of concept. While the pathways for retinal gliosis and progressive tissue degeneration are well-documented in these models, integrating high-dimensional transcriptomics may provide deeper functional insights.

The analysis suggests that it is feasible to explore potential regulatory components and identify candidate IHC targets using a standard 16 GB laptop and publicly available RNA-seq data. Applying appropriate memory management and garbage collection routines can help maintain stability within these hardware constraints. Furthermore, exploring aggregated databases allows for broader cross-referencing, which might help reduce reliance on established markers and offer alternative, data-driven avenues for discovering and validating new targets in complex retinopathies.
